SMT-Boosted Security Types and Correctness for Low-Level MPC

### 2025-11-24 toplas26/frontmatter.tex
checkpoint

\title{SMT-Boosted Security Types for Low-Level MPC}

@@

\begin{abstract}
  Secure Multi-Party Computation (MPC) is an important enabling
  technology for data privacy in modern distributed applications. We
  develop a new type theory to automatically enforce correctness,
  confidentiality, and integrity properties of protocols written in
  the \emph{Prelude/Overture} language framework. Judgements in the
  type theory are predicated on SMT verifications in a theory of
  finite fields, which supports precise and efficient analysis. Our
  approach is automated, compositional, scalable, and generalizes to
  arbitrary prime fields.
\end{abstract}

>>>>

\title{SMT-Boosted Security Types and Correctness for Low-Level MPC}

@@

\begin{abstract}
  Secure Multi-Party Computation (MPC) is an important enabling
  technology for data privacy in modern distributed applications. We
  develop a new Hoare logic and type theory to automatically enforce
  correctness, confidentiality, and integrity properties of protocols
  written in the \emph{Prelude/Overture} language framework. Both
  definition and automation of our analyses leverages SMT verification
  in a theory of finite fields, supporting precision and
  efficiency. Our approach is compositional, scalable, and generalizes
  to arbitrary prime fields and secret sharing schemes.
\end{abstract}

### 2025-11-24 toplas26/intro.tex
checkpoint

Data privacy is a critical concern in distributed applications
including privacy-preserving machine learning
\cite{li2021privacy,knott2021crypten,koch2020privacy,liu2020privacy}
and blockchains
\cite{ishai2009zero,lu2019honeybadgermpc,gao2022symmeproof,tomaz2020preserving}.
Secure multiparty computation (MPC) is a key software-based enabling
technology for these applications. MPC protocols provide both confidentiality
and integrity properties, formulated as real/ideal (aka simulator)
security and universal composability (UC) \cite{evans2018pragmatic},
with well-studied manual proof methods \cite{Lindell2017}.  MPC
security semantics has also been reformulated as
\emph{hyperproperties}
\cite{8429300,10.1145/3453483.3454074,skalka-near-ppdp24} to support
automated language-based proof methods. Recent research in the SMT
community has developed theories of finite fields \cite{SMFF} with
clear relevance to verification of cryptographic schemes that rely on
field arithmetic. The goal of this paper is to combine these two
approaches in a decidable type system for verifying correctness and
security properties of MPC protocols.

>>>>

Data privacy is a critical concern in distributed applications
including privacy-preserving machine learning
\cite{li2021privacy,knott2021crypten,koch2020privacy,liu2020privacy}
and blockchains
\cite{ishai2009zero,lu2019honeybadgermpc,gao2022symmeproof,tomaz2020preserving}.
Secure multiparty computation (MPC) is a key software-based enabling
technology for these applications. MPC protocols provide both confidentiality
and integrity properties, formulated as real/ideal (aka simulator)
security and universal composability (UC) \cite{evans2018pragmatic},
with well-studied manual proof methods \cite{Lindell2017}.  MPC
security semantics has also been reformulated as
\emph{hyperproperties}
\cite{8429300,10.1145/3453483.3454074,skalka-near-ppdp24} to support
automated language-based proof methods. Recent research in the SMT
community has developed theories of finite fields \cite{SMFF} with
clear relevance to verification of cryptographic schemes that rely on
field arithmetic. The goal of this paper is to combine hyperproperty
formulations of security with SMT verification to enforce correctness
and security properties of MPC protocols.

### 2025-11-24 toplas26/sty.tex
checkpoint

Our confidentiality analysis mainly considers protocols in terms of
their SMT interpretations as described in Section \ref{section-smt}.
We introduce a \emph{share type} language $\sty$, classifying
\emph{share terms}. We represent share terms using standard notation
\cite{10.1007/978-3-030-68869-1_3}. A share term $[\phi_1,\phi_2]$
represents a secret-shared value where client 1 has $\phi_1$ and
client 2 has $\phi_2$, and this shared value can be reconstructed
as a public value $\pubx{\phi}$.
$$
\begin{array}{rcl@{\hspace{10mm}}l}
  \sty &::=& \sharety \mid \pubty \mid \outty & \textit{share types} \\
  \stt &::=& [\phi,\phi] \mid \pubx{\phi} & \textit{share terms} 
\end{array}
$$
\paragraph{Share generation and reconstruction}
A \emph{secret sharing scheme} includes functions
$(\genshares,\recon)$ that generate and reconstruct shares,
respectively. The $\genshares$ function takes some
input $\phi$ and randomness $\rx{x}{w}$ and produces
two shares (or as many is needed for the number of clients
in the federation): 
$$
\genshares(\phi,\rx{w}{\cid}) = \phi_1,\phi_2
$$
and:
$$
\recon(\genshares(\phi,\rx{w}{\cid})) = \phi
$$
In order for $\genshares$ to be \emph{sound}, we require
that both of $\phi_1$ and $\phi_2$ are marginally independent
of $\phi$, and also of any other values in the protocol.
This entails the requirement that $\rx{w}{\cid}$ is
consumed by $\genshares$, e.g., as an OTP.
We formally characterize soundness of $\genshares$
in our protocol model in Definition \ref{def-genshares}.
We emphasize that our type system is agnostic with respect
to the definition of $\genshares$ and $\recon$ modulo
soundness-- we use an additive shares encoding in this work,
but the analysis could apply instead to, e.g., Shamir
sharing.

@@

Our non-interactive typing rules are presented in Figure
\ref{fig-stylocal}. Judgmements are of the form $\Gamma, \eqs \vdash
\stt : \sty$, where $\Gamma$ is a set of \emph{share type bindings} of
the form $\stt : \sty$. In the context of typing a protocol $\prog$
soundness requires that the constraint $\eqs$ must be entailed by
$\eqspre \wedge \toeq{\prog}$ (see, e.g., Lemma
\ref{lemma-sty-noninteractive-sound}).

The syntax-directed rules are all known semi-homomorphic operations,
including addition of shares (\TirName{HEAdd}), and both addition and
multiplication of shares with a public value (\TirName{HEAddPub} and
\TirName{HEMultPub}, respectively) \cite{10.1007/978-3-030-68869-1_3}.
These rules also leverage SMT entailment in the \TirName{ShareEntails}
and \TirName{PubEntails} rules, since these operations should be
allowable up to algebraic equivalence in the context of a protocol.

@@

The \TirName{GenShares} rule generates the binding $[x_1,
  x_2] : \sharety$ if the term is a secret sharing as defined
by $\genshares$, and adds $\rx{y}{\cid_1}$ to the set of variables
consumed by $\genshares$, ensuring one-time usage. The
\TirName{Reveal} rule generates the binding $\pubx{\recon(\px{x},\px{y})} :
\outty$ with the requirement that $[\px{x},\px{y}]$ are typable as
shares. The \TirName{SecOpen} rule is similar except it allows the
assertion $\pubx{\recon(\px{x},\px{y})} : \pubty$ if we can additionally
show that the shared value is masked by the use of random value
$\rx{z}{\oid}$ as an OTP. Here we are positing a third party $\oid$
that secret-shares $\rx{z}{\oid}$ as part of an initialization
protocol. We illustrate secure openings in practice at more length
in Section \ref{section-examples}, though we note that appropriate
properties of initialization protocols can be encoded in $\eqspre$.

@@

It is well-known that adding or subtracting a sample from a uniform
distribution in a finite field yields a value in a uniform
distribution, meaning that samples can be used as OTPs with
perfect secrecy \cite{barthe2019probabilistic,darais2019language}.
So we can define:
\begin{eqnarray*}
  \genshares(\phi,\rx{y}{\cid}) &\defeq& \phi - \rx{y}{\cid}, \rx{y}{\cid}\\
  \recon(\phi_1,\phi_2) &\defeq& \phi_1 \fplus \phi_2
\end{eqnarray*}
where we require that $\genshares$ consumes the randomness
$\setit{\rx{y}{\cid_2}}$ for use as an OTP.

This approach highlights an imporant interaction between our type
system and SMT entailment as defined by the $\TirName{ShareEntails}$
rule. In particular there are other methods for encrypting values in MPC
protocols that can be shown to be algebraically equivalent
to this fundamental schema. For example, in Yao's Garbled Circuits
(YGC) in $\mathbb{F}_{2}$, the ``garbler'' represents encrypted wire
values as a random sample (for 1) or its negation (for 0), and shares
its secret input encoding with the ``evaluator'' by using
it as a selection bit. That is, assuming that client 2 is the garbler and
client 1 is the evaluator, we can define:
$$
\xassign{\mx{w}{1}}{\mux{\secret{w}}{\flip{w}}{\neg\flip{w}}}{2}
$$
where for all $\be_1,\be_2,\be_3$ (with $\neg$ denoting negation in $\mathbb{F}_2$):
$$
\mux{\be_1}{\be_2}{\be_3} \defeq (\neg\be_1 \ftimes \be_2) \fplus (\be_1 \ftimes \be_3)
$$
and letting this protocol be $\prog$ the following is valid:
\begin{mathpar}
%\toeq{\xassign{\mx{w}{1}}{\OT{\secret{w}}{1}{\neg\flip{w}}{\flip{w}}}{2}}
   \toeq{\prog} \models \mx{w}{1} \eop \sx{w}{2} \fminus \rx{w}{2}
\end{mathpar}
For this reason SMT entailment is also an important enabling technology
for the interactive rules.

\subsection{Share Typing Example} To illustrate details of the
type system we return to the example of additive sharing in the context
of a 2-client protocol. We first focus on initial secret sharing steps--
let $\prog$ be the following protocol in $\fieldp{2}$ (noting
that $-$ is equivalent to xor in $\fieldp{2}$):
$$
\begin{array}{lll}
  \elab{\mesg{s1}}{2} := \elab{(\secret{1} \fminus \locflip)}{1}; \quad
  \elab{\mesg{s2}}{1} := \elab{(\secret{2} \fminus \locflip)}{2} 
\end{array}
$$
By $\TirName{GenShares}$ we can then assert:
$$
\toeq{\prog} \vdash \varnothing,\varnothing \stredx
\setit{[\mx{s2}{1}, \elab{\locflip}{2}] : \sharety,
  [\elab{\locflip}{1}, \mx{s1}{2}] : \sharety}, \setit{\elab{\locflip}{1},\elab{\locflip}{2}}
$$
Suppose the protocol goes on to reveal the negation of the
sum (xor-ing) of the inputs:
$$
\begin{array}{l}
  \px{1} := \elab{\mesg{s2} + \locflip + 1}{1};\quad \px{2} := \elab{\mesg{s1} + \locflip}{2};\\
  \out{1} := \elab{\px{1} + \px{2}}{1};\quad \out{2} := \elab{\px{1} + \px{2}}{2}
\end{array}
$$
Letting this extended protocol be $\prog'$, and letting:
$$
\begin{array}{rcl}
\Gamma &\defeq& \setit{[\mx{s2}{1}, \elab{\locflip}{2}] : \sharety,
  [\elab{\locflip}{1}, \mx{s1}{2}] : \sharety}\\
R &\defeq&  \setit{\elab{\locflip}{1},\elab{\locflip}{2}}
\end{array}
$$
observe we can then obtain share typings for noninteractive components for
addition of shares by \TirName{Shared} and \TirName{HeAdd}:
$$
\Gamma, \toeq{\prog'} \vdash
[\mx{s2}{1} + \elab{\locflip}{1}, \mx{s1}{2} + \elab{\locflip}{2}] : \sharety
$$
and hence for negation by \TirName{HeAddPub}:
$$
\Gamma, \toeq{\prog'} \vdash
[\mx{s2}{1} + \elab{\locflip}{1} + 1, \mx{s1}{2} + \elab{\locflip}{2}] : \sharety
$$
This implies:
$$
\toeq{\prog'} \vdash \varnothing,\varnothing \stredx
\Gamma \cup \setit{\pubx{\px{1} + \px{2}} : \outty}, R 
$$
Hence by \TirName{PubEntails} we have:
$$
\begin{array}{l}
\Gamma \cup \setit{\pubx{\px{1} + \px{2}} : \outty}, \toeq{\prog'} \vdash
\pubx{\out{1}} : \outty \\
\Gamma \cup \setit{\pubx{\px{1} + \px{2}} : \outty}, \toeq{\prog'} \vdash
\pubx{\out{2}} : \outty
\end{array}
$$

@@

guarantees confidentiality (Theorem \ref{theorem-OK}).

>>>>

Our confidentiality analysis mainly considers protocols in terms of
their SMT interpretations as described in Section \ref{section-smt}.
We introduce a \emph{share type} language $\sty$, classifying
\emph{share terms}. We represent share terms using standard notation
\cite{10.1007/978-3-030-68869-1_3}. A share term $[\phi_1,\phi_2]$
represents a secret-shared value where client 1 has $\phi_1$ and
client 2 has $\phi_2$, and this shared value can be reconstructed
as a public value $\pubx{\phi}$.
$$
\begin{array}{rcl@{\hspace{10mm}}l}
  \sty &::=& \sharety \mid \pubty \mid \outty & \textit{share types} \\
  \stt &::=& [\phi,\phi] \mid \pubx{\phi} & \textit{share terms} 
\end{array}
$$
\paragraph{Share generation and reconstruction}
A \emph{secret sharing scheme} includes functions
$(\genshares,\recon)$ that generate and reconstruct shares,
respectively. The $\genshares$ function takes some
input $\phi$ and randomness $\rx{x}{w}$ and produces
two shares (or as many is needed for the number of clients
in the federation): 
$$
\genshares(\phi,\rx{w}{\cid}) = \phi_1,\phi_2
$$
and:
$$
\recon(\genshares(\phi,\rx{w}{\cid})) = \phi
$$
In order for $\genshares$ to be \emph{sound} with regard to
confidentiality, we require
that both of $\phi_1$ and $\phi_2$ are marginally independent
of $\phi$, and also of any other values in the protocol.
This entails the requirement that $\rx{w}{\cid}$ is
consumed by $\genshares$, e.g., as an OTP.
We formally characterize soundness of $\genshares$
in our protocol model in Definition \ref{def-genshares}.
We emphasize that our type system is agnostic with respect
to the definition of $\genshares$ and $\recon$ modulo
soundness-- we use an additive shares encoding in this work,
but the analysis could apply instead to, e.g., Shamir
sharing.

@@

Our non-interactive typing rules are presented in Figure
\ref{fig-stylocal}. Judgmements are of the form $\Gamma, \eqs \vdash
\stt : \sty$, where $\Gamma$ is a set of \emph{share type bindings} of
the form $\stt : \sty$. In the context of typing a protocol $\prog$
soundness requires that the constraint $\eqs$ must be entailed by
$\toeq{\prog}$. The syntax-directed rules are all known semi-homomorphic operations,
including addition of shares (\TirName{HEAdd}), and both addition and
multiplication of shares with a public value (\TirName{HEAddPub} and
\TirName{HEMultPub}, respectively) \cite{10.1007/978-3-030-68869-1_3}.
These rules also leverage SMT entailment in the \TirName{ShareEntails}
and \TirName{PubEntails} rules, meaning these operations are
allowable up to algebraic equivalence in the context of a protocol.

@@

The \TirName{GenShares} rule generates the binding $[x_1,
  x_2] : \sharety$ if the term is a secret sharing as defined
by $\genshares$, and adds $\rx{w}{\cid}$ to the set of variables
consumed by $\genshares$, ensuring one-time usage. The
\TirName{Reveal} rule generates the binding $\pubx{\recon(\px{w_1},\px{w_2})} :
\outty$ with the requirement that $[\px{w_1},\px{w_2}]$ are typable as
shares. The \TirName{SecOpen} rule is similar except it allows the
assertion $\pubx{\recon(\px{w_1},\px{w_2})} : \pubty$ if we can additionally
show that the shared value is masked by the use of random value
$\rx{w}{\oid}$ as an OTP. Here we are positing a third party $\oid$
that secret-shares $\rx{w}{\oid}$ as part of an initialization
protocol. We illustrate secure openings in practice at more length
in Section \ref{section-examples}, though we note that appropriate
properties of initialization protocols can be encoded in $\eqspre$.

@@

It is well-known that adding or subtracting a sample from a uniform
distribution in a finite field yields a value in a uniform
distribution, meaning that samples can be used as OTPs with
perfect secrecy \cite{barthe2019probabilistic,darais2019language}.
So we can define:
\begin{eqnarray*}
  \genshares(\phi,\rx{w}{\cid}) &\defeq& \phi - \rx{w}{\cid}, \rx{w}{\cid}\\
  \recon(\phi_1,\phi_2) &\defeq& \phi_1 \fplus \phi_2
\end{eqnarray*}
where we require that $\genshares$ consumes the random value
$\rx{w}{\cid}$ as an OTP.

This approach highlights an important interaction between our type
system and SMT entailment as defined by the $\TirName{ShareEntails}$
rule. In particular there are other methods for encrypting values in MPC
protocols that can be shown to be algebraically equivalent
to this fundamental schema. For example, in Yao's Garbled Circuits
(YGC) in $\mathbb{F}_{2}$, the ``garbler'' represents encrypted wire
values as a random sample (for 0) or its negation (for 1), and shares
its secret input encoding with the ``evaluator'' by using
it as a selection bit. That is, assuming that client 2 is the garbler and
client 1 is the evaluator, we can define:
$$
\xassign{\mx{w}{1}}{\mux{\secret{w}}{\flip{w}}{\neg\flip{w}}}{2}
$$
where for all $\be_1,\be_2,\be_3$ (with $\neg$ denoting negation in $\mathbb{F}_2$):
$$
\mux{\be_1}{\be_2}{\be_3} \defeq (\neg\be_1 \ftimes \be_2) \fplus (\be_1 \ftimes \be_3)
$$
and letting this protocol be $\prog$ the following is valid:
\begin{mathpar}
%\toeq{\xassign{\mx{w}{1}}{\OT{\secret{w}}{1}{\neg\flip{w}}{\flip{w}}}{2}}
   \toeq{\prog} \models \mx{w}{1} \eop \sx{w}{2} \fminus \rx{w}{2}
\end{mathpar}
For this reason SMT entailment is also an important enabling technology
for the interactive rules.

\subsection{Share Typing Example} To illustrate details of the
type system we return to the example of additive sharing in the context
of a 2-client protocol. While the ideal functionality of this simple protocol
precisely reveals input secrets, it serves as an understandable example
of the type system mechanics.
We first focus on initial secret sharing steps--
let $\prog$ be the following protocol in $\fieldp{2}$ (noting
that $-$ is equivalent to xor in $\fieldp{2}$):
$$
\begin{array}{lll}
  \elab{\mesg{s1}}{2} := \elab{(\secret{1} \fminus \locflip)}{1}; \quad
  \elab{\mesg{s2}}{1} := \elab{(\secret{2} \fminus \locflip)}{2} 
\end{array}
$$
By $\TirName{GenShares}$ we can then assert:
$$
\toeq{\prog} \vdash \varnothing,\varnothing \stredx
\setit{[\mx{s2}{1}, \elab{\locflip}{2}] : \sharety,
  [\elab{\locflip}{1}, \mx{s1}{2}] : \sharety}, \setit{\elab{\locflip}{1},\elab{\locflip}{2}}
$$
Suppose the protocol goes on to reveal the negation of the
sum (xor-ing) of the inputs:
$$
\begin{array}{l}
  \px{1} := \elab{\mesg{s2} + \locflip + 1}{1};\quad \px{2} := \elab{\mesg{s1} + \locflip}{2};\\
  \out{1} := \elab{\px{1} + \px{2}}{1};\quad \out{2} := \elab{\px{1} + \px{2}}{2}
\end{array}
$$
Letting this extended protocol be $\prog'$, and letting:
$$
\begin{array}{rcl}
\Gamma &\defeq& \setit{[\mx{s2}{1}, \elab{\locflip}{2}] : \sharety,
  [\elab{\locflip}{1}, \mx{s1}{2}] : \sharety}\\
R &\defeq&  \setit{\elab{\locflip}{1},\elab{\locflip}{2}}
\end{array}
$$
observe we can then obtain share typings for noninteractive components for
addition of shares by \TirName{Shared} and \TirName{HeAdd}:
$$
\Gamma, \toeq{\prog'} \vdash
[\mx{s2}{1} + \elab{\locflip}{1}, \mx{s1}{2} + \elab{\locflip}{2}] : \sharety
$$
and hence for negation by \TirName{HeAddPub}:
$$
\Gamma, \toeq{\prog'} \vdash
[\mx{s2}{1} + \elab{\locflip}{1} + 1, \mx{s1}{2} + \elab{\locflip}{2}] : \sharety
$$
This implies:
$$
\toeq{\prog'} \vdash \varnothing,\varnothing \stredx
\Gamma \cup \setit{\pubx{\px{1} + \px{2}} : \outty}, R 
$$
Hence by \TirName{PubEntails} we have:
$$
\begin{array}{l}
\Gamma \cup \setit{\pubx{\px{1} + \px{2}} : \outty}, \toeq{\prog'} \vdash
\pubx{\out{1}} : \outty \\
\Gamma \cup \setit{\pubx{\px{1} + \px{2}} : \outty}, \toeq{\prog'} \vdash
\pubx{\out{2}} : \outty
\end{array}
$$

@@

the invariant guarantees confidentiality (Theorem \ref{theorem-OK}).
